regular languages

### 2023-12-05 dfa.tex
edit

\begin{defn}
  The language of a regular expression $r$ is defined by recursion on $r$:
  \begin{enumerate}
  \item the language of the regular expression $\varnothing$ is the empty language:
    $$\sL(\varnothing) = \varnothing$$
  \item the language of the regular expression $\lambda$ is the language containing just the empty string:
    $$\sL(\varepsilon) = \Set{\varepsilon}$$
  \item the language of the regular expression $c$ is the language containing just $c$: for each $c \in \Sigma$,
    $$\sL(c) = \Set{c}$$
  \item the language of the regular expression $rr'$ is the concatenation of $\sL(r)$ and $\sL(r')$:
    $$\sL(r \cdot r') = \sL(r) \cdot \sL(r')$$
  \item the language of the regular expression $r + r'$ is the union of $\sL(r)$ and $\sL(r')$:
    $$\sL(r + r') = \sL(r) \cup \sL(r')$$
  \item the language of the regular expression $r^{*}$ is the Kleene closure of $\sL(r)$:
    $$\sL(r^{*}) = (\sL(r))^{*}$$
  \end{enumerate}
\end{defn}

\begin{thm}\label{thm:regex-to-dfa}
  Let $e$ be a regular expression.
  $\sL(e)$ is regular, i.e., there is a DFA $D$ that recognizes $\sL(e)$.
\end{thm}
\begin{proof}
  \begin{node}
    We proceed by induction on the structure of $e$.
    When $e$ is $\varnothing, \varepsilon,$ or $c$, $\sL(e)$ is finite.
    In \cref{eg:finite-languages-regular}, we informally argued that finite languages are regular.
  \end{node}
  \begin{node}
    When $e$ is $r \cdot r'$ or $r + r'$, the induction hypothesis states that $\sL(r)$ and $\sL(r')$ are regular.
    By definition $\sL(r \cdot r') = \sL(r) \cdot \sL(r')$ and $\sL(r + r') = \sL(r) \cup \sL(r')$.
    By \cref{thm:regular-closure}, the class of regular languages is closed under concatenation and union, so $\sL(r \cdot r')$ and $\sL(r + r')$ are regular.
  \end{node}
  \begin{node}
    Finally, when $e$ is $r^{*}$, the induction hypothesis again states that $\sL(r)$ is regular.
    By definition $\sL(r^{*}) = (\sL(r))^{*}$.
    By \cref{thm:regular-closure}, the class of regular languages is closed under Kleene closure, so $\sL(r^{*})$ is regular.
  \end{node}
\end{proof}

\begin{node}
  \cref{thm:regex-to-dfa} shows that every regular expression has a corresponding DFA.
  The converse is also true: every DFA has a corresponding regular expression.
  It's proof is less trivial.
\end{node}

>>>>

\begin{defn}
  The language of a regular expression $r$ is defined by recursion on $r$:
  \begin{enumerate}
  \item the language of the regular expression $\varnothing$ is the empty language:
    $$\sL(\varnothing) = \varnothing$$
  \item the language of the regular expression $\varepsilon$ is the language containing just the empty string:
    $$\sL(\varepsilon) = \Set{\varepsilon}$$
  \item the language of the regular expression $c$ is the language containing just $c$: for each $c \in \Sigma$,
    $$\sL(c) = \Set{c}$$
  \item the language of the regular expression $rr'$ is the concatenation of $\sL(r)$ and $\sL(r')$:
    $$\sL(r \cdot r') = \sL(r) \cdot \sL(r')$$
  \item the language of the regular expression $r + r'$ is the union of $\sL(r)$ and $\sL(r')$:
    $$\sL(r + r') = \sL(r) \cup \sL(r')$$
  \item the language of the regular expression $r^{*}$ is the Kleene closure of $\sL(r)$:
    $$\sL(r^{*}) = (\sL(r))^{*}$$
  \end{enumerate}
\end{defn}

\begin{thm}\label{thm:regex-to-dfa}
  Let $e$ be a regular expression.
  $\sL(e)$ is regular, i.e., there is a DFA $D$ that recognizes $\sL(e)$.
\end{thm}
\begin{proof}
  \begin{node}
    We proceed by induction on the structure of $e$.
    When $e$ is $\varnothing, \varepsilon,$ or $c$, $\sL(e)$ is finite.
    In \cref{eg:finite-languages-regular}, we informally argued that finite languages are regular.
  \end{node}
  \begin{node}
    When $e$ is $r \cdot r'$ or $r + r'$, the induction hypothesis states that $\sL(r)$ and $\sL(r')$ are regular.
    By definition $\sL(r \cdot r') = \sL(r) \cdot \sL(r')$ and $\sL(r + r') = \sL(r) \cup \sL(r')$.
    By \cref{thm:regular-closure}, the class of regular languages is closed under concatenation and union, so $\sL(r \cdot r')$ and $\sL(r + r')$ are regular.
  \end{node}
  \begin{node}
    Finally, when $e$ is $r^{*}$, the induction hypothesis again states that $\sL(r)$ is regular.
    By definition, $\sL(r^{*}) = (\sL(r))^{*}$.
    By \cref{thm:regular-closure}, the class of regular languages is closed under Kleene closure, so $\sL(r^{*})$ is regular.
  \end{node}
\end{proof}

\begin{node}
  \cref{thm:regex-to-dfa} shows that every regular expression has a corresponding DFA.
  The converse is also true: every DFA has a corresponding regular expression.
  The proof of this fact is not trivial.
\end{node}
